---
title: "P8105 Homework 2"
author: "By Wenda Yoanda, UNI:wy2486" 
output: github_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1: NYC Transit data

**Read and clean the data**
```{r message=FALSE, warning=FALSE}
nycsubway_df =
  read_csv(
    file = "./datahw2/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("", "NA", ".")
    )
  
nycsubway_df = janitor::clean_names(nycsubway_df)
```

**Retain relevant variables and recode *entry* **
```{r message=FALSE, warning=FALSE}
nycsubwayfinal_df =
nycsubway_df |>
select(line, 
       station_name, 
       station_latitude, 
       station_longitude, 
       starts_with("route"),
       entry,
       vending,
       entrance_type,
       ada) |>
  mutate(
    entry= case_match(
      entry,
      "YES"~ TRUE,
      "NO"~ FALSE)
    )
```

The original dataset `nycsubway_df` contains 1,868 observations and 32 variables. The dataset was cleaned by:

1. Standardizing variable names to lowercase with underscores using `clean_names()`.
2. Treating blank values, `"NA"`, and `"."` as missing values.
3. Retaining only the variables relevant to the analysis using `select()`.
4. Converting the `entry` variable from `"YES"` / `"NO"` to a logical variable (`TRUE` / `FALSE`) using `mutate()` and `case_match()`.

After these steps, the cleaned dataset `nycsubwayfinal_df` contains 1,868 observations and 19 variables, resulting in a dimension of 1,868 × 19. The retained variables include *Line, Station Name, Station Latitude, Station Longitude, Routes (1–11), Entry, Vending, Entrance Type,* and *ADA Compliance*. The dataset is not fully tidy because the routes served are stored across multiple columns rather than in a single route variable.


















