---
title: "P8105 Homework 2"
author: "By Wenda Yoanda, UNI:wy2486" 
output: github_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1: NYC Transit data

**Read and clean the dataset**
```{r message=FALSE, warning=FALSE}
nycsubway_df =
  read_csv(
    file = "./datahw2/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("", "NA", ".")
    )
  
nycsubway_df = janitor::clean_names(nycsubway_df)
```

**Retain relevant variables and recode *entry* **
```{r message=FALSE, warning=FALSE}
nycsubwayfinal_df =
nycsubway_df |>
select(line, 
       station_name, 
       station_latitude, 
       station_longitude, 
       starts_with("route"),
       entry,
       vending,
       entrance_type,
       ada) |>
  mutate(
    entry= case_match(
      entry,
      "YES"~ TRUE,
      "NO"~ FALSE)
    )
```

The original dataset `nycsubway_df` contains 1,868 observations and 32 variables. The dataset was cleaned by:

1. Standardizing variable names to lowercase with underscores using `clean_names()`.
2. Treating blank values, `"NA"`, and `"."` as missing values.
3. Retaining only the variables relevant to the analysis using `select()`.
4. Converting the `entry` variable from `"YES"` / `"NO"` to a logical variable (`TRUE` / `FALSE`) using `mutate()` and `case_match()`.

After these steps, the cleaned dataset `nycsubwayfinal_df` contains 1,868 observations and 19 variables, resulting in a dimension of 1,868 × 19. The retained variables include *Line, Station Name, Station Latitude, Station Longitude, Routes (1–11), Entry, Vending, Entrance Type,* and *ADA Compliance*. The dataset is not fully tidy because the routes served are stored across multiple columns rather than in a single route variable.

### Analyzing The Cleaned Dataset
**How many distinct stations are there?**
```{r echo=FALSE, message=FALSE, warning=FALSE}
nrow(
  distinct(
    nycsubwayfinal_df,
    station_name,
    line
  )
)
```
**How many stations are ADA compliant?**
```{r echo=FALSE, message=FALSE, warning=FALSE}
nycsubwayfinal_df |>
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```
**What proportion of station entrances / exits without vending allow entrance (%)?**
```{r echo=FALSE, message=FALSE, warning=FALSE}
round(
  nrow(
    filter(
      nycsubwayfinal_df,
      vending == "NO",
      entry == TRUE
    )
  ) /
  nrow(
    filter(
      nycsubwayfinal_df,
      vending == "NO"
    )
  ) * 100,
  2
)
```
**Reformatting Data** 

Reformat so that route number and route name are distinct variables
<!-- I use ChatGPT here to debug my code, as there is error message saying that there are some of the route datasets that are not all the same type of data, so needed to be changed to character, hence, I use the mutate(across(route1:route11, as.character) -->

```{r echo=FALSE, message=FALSE, warning=FALSE}
nycsubwayreformat_df =
  nycsubwayfinal_df |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )

nycsubwayreformat_df |>
  head(10) |>
  knitr::kable()
```

**How many distinct stations serve the A train?**
```{r echo=FALSE, message=FALSE, warning=FALSE}
nycsubwayreformat_df |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()
```

**Of the stations that serve the A train, how many are ADA compliant?**
```{r echo=FALSE, message=FALSE, warning=FALSE}
nycsubwayreformat_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(station_name,line) |>
  nrow()
```
## Problem 2: Trash Wheel Dataset

**Read and clean the *Mr. Trash Wheel* dataset**
```{r message=FALSE, warning=FALSE}
mr_trash_wheel_df =
  read_excel(
    "./datahw2/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742",
    na = c("", "NA", ".")
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )
```

**Read and clean the *Professor Trash Wheel* dataset**
```{r message=FALSE, warning=FALSE}
prof_trash_wheel_df =
  read_excel(
    "./datahw2/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    na = c("", "NA", ".")
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )
```

**Read and clean the *Gwynnda* dataset**
```{r message=FALSE, warning=FALSE}
gwynnda_df =
  read_excel(
    "./datahw2/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394",
    na = c("", "NA", ".")
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Gwynnda"
  )
```

**Combining All Datasets**

Combine *Mr. Trash Wheel*, *Professor Trash Wheel* and *Gwynnda* Datasets
```{r message=FALSE, warning=FALSE}
alltrashwheel_df =
  bind_rows(
    mr_trash_wheel_df,
    prof_trash_wheel_df,
    gwynnda_df
  )
```
A snapshot of the first 10 observations in the final combined dataset is shown below:
```{r echo=FALSE, message=FALSE, warning=FALSE}
alltrashwheel_df |>
  head(10) |>
  knitr::kable()
```

The combined `alltrashwheel_df` dataset contains `r nrow(alltrashwheel_df)` observations and `r ncol(alltrashwheel_df)` variables. Key variables include *dumpster*, *month*, *year*, *date*, *weight_tons*, *volume_cubic_yards*, *plastic_bottles*, *polystyrene*, *cigarette_butts*, *glass_bottles*, *plastic_bags*, *wrappers*, *sports_balls*, *homes_powered*, and *trash_wheel*. The three Trash Wheel datasets were imported from separate Excel sheets, cleaned using standardized variable names, restricted to dumpster-specific observations, and combined using `bind_rows()`. 

During the cleaning process, variables such as *year* and *plastic_bottles* were converted to consistent integer or numeric formats because their data types were not standardized across the three datasets. A *trash_wheel* variable was also added before combining the datasets to identify the source of each observation.

The total weight of trash collected by Professor Trash Wheel is **`r alltrashwheel_df |> filter(trash_wheel == "Professor Trash Wheel") |> summarize(total = sum(weight_tons, na.rm = TRUE)) |> pull(total)`** tons.

The total number of cigarette butts collected by Gwynnda in June of 2022 is **`r alltrashwheel_df |> filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |> summarize(total = sum(cigarette_butts, na.rm = TRUE)) |> pull(total)`**.















